---
title: "class11"
author: "Eli Sobel A69027989"
format: pdf
---

Today, before delving into structure prediction with AlphaFold we will finish off our previous lab10 "comparative structure analysis" section.

```{r}
library(bio3d)

id <- "1ake_A"

aa <- get.seq(id)
aa
```

```{r}
b <- blast.pdb(aa)
```
```{r}
attributes(b)
```

```{r}
head(b$hit.tbl)
```

```{r}
hits <- plot(b)
```
```{r}
attributes(hits)
```
Top hits that we like from our blast results:
```{r}
hits$pdb.id
```

# find and download all related structures

```{r}
# donwload related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

I have now found and downloaded all ADK structures in the PDB database but viewing them is difficult as they need to be aligned and superposed.

I am going to install BiocManager package from CRAN. Then I can use "BiocManager::install("msa")".

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```



## Principal Component analysis

```{r}
pc <- pca(pdbs)
plot(pc)
```

```{r}
plot(pc, pc.axes = c(1:2))
```

To examine in more detail what PC1 (or any PC) is capturing here we an plot the loadings or make a movie (trajectory) of moving along PC1

```{r}
mktrj(pc, pc=1, file="pc1.pdb")
```


