---
title: "class 12 homework"
author: "Eli Sobel A69027989"
format: pdf
---

## Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.
This is the final file you got (https://bioboot.github.io/bggn213_F24/class-material/rs8067378_ENSG00000172057.6.txt). The first column is sample name, the second column is genotype and the third column are the expression values.
Open a new RMarkdown document in RStudio to answer the following two questions. Submit your resulting PDF report with your working code, output and narrative text answering Q13 and Q14 to GradeScope.

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

The sample size is 108 for A/A, 233 for A/G, and 121 for G/G. The median expression levels per genotype are 31.25 for A/A, 25.06 for A/G, and 20.07 for G/G. Code provided below.

Read in data and examine the structure with `head()`:

```{r} 
expr <- read.table("https://bioboot.github.io/bggn213_F24/class-material/rs8067378_ENSG00000172057.6.txt")
head(expr)
```

Generate table of samples per genotype with `table()`:

```{r}
table(expr$geno)
```

Calculate median expression per genotype:

```{r}
# make sure tidyr and dplyr are installed via install.packages()
library(tidyr)
library(dplyr)

median_exp_per_geno <- expr %>%
  group_by(geno) %>%
  summarise(median_exp = median(exp))
median_exp_per_geno
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

You could infer from the relatively higher expression range of A/A compared to G/G, as well as the intermediate range of A/G, that the G->A SNP is associated with increased expression of ORMDL3. It is likely that there is a causal relationship between this SNP and ORMDL3 expression.

Use ggplot to make a box plot of expression by genotype:

```{r}
library(ggplot2)

x <- ggplot(expr, aes(geno, exp, col=geno)) + geom_boxplot() + labs(x = "genotype", y = "expression (FPKM)")
x
```



